\Title

### 2023-07-04 Grimaldi-etal-BiolCybernetics.tex
Update Grimaldi-etal-BiolCybernetics.tex

The precise timing of spikes from afferent neurons determines the response of a biological neuron. This temporal aspect of neural activity is essential for understanding information processing in neurobiology, and is particularly applicable to the output of neuromorphic hardware such as event-based cameras. However, most artificial neural models ignore this important temporal dimension of neural activity. Here, we develop a model for efficiently detecting temporal spiking motifs based on a layer of spiking neurons with heterogeneous synaptic delays. The model exploits the property that the diversity of synaptic delays on the dendritic tree allows specific arrangements of temporally precise synaptic inputs to synchronize as they reach the basal dendritic tree. We show that this process can be formalized as a time-invariant logistic regression that can be trained on labeled data. We demonstrate its application to naturalistic videos transformed into event streams, similar to the output of the biological retina or event-based cameras. To characterize the robustness of the model in detecting visual motion, we test its efficiency by pruning some weights and show that it is still efficient at very low workloads. In conclusion, heterogeneous delays provide an end-to-end, event-driven computational building block that could improve the performance of future spiking neural network (SNN) algorithms and, in particular, their applications in neuromorphic chips.

@@

\keywords{\Keywords}
%
\maketitle
%
\section{Introduction}
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable capacity to react efficiently at any given time while consuming a reasonable amount of energy, in the order of tens of watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation that both use. For instance, the convolutional neural networks (CNNs) which are used for processing images represent the flow of information from one layer to another using tensors of floats, storing visual information densely across the visual topography, with different translation-invariant properties represented in different channels. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. These networks are also known to mimic several properties of biological systems, such that each can be assigned a ``brain score''~\citep{schrimpf_brain-score_2020}. However, this score does not take into account key aspects of the efficiency of biological systems, such as inference speed or energy consumption. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing (binary) events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. Spiking neural networks (SNNs), known as the third generation of artificial neural networks, incorporate this temporal dimension into the way they perform their computations. One example of a SNN is the SpikeNet algorithm, which takes a purely temporal approach by encoding information using at most one spike per neuron~\citep{delorme_spikenet_1999}. Alternatively, the SNN implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} that uses the exact timing of spikes and learns the structure of the network by minimizing a specifically defined cost function. This was recently extended using the surrogate gradient method which is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of networks using an analog, firing rate-based representation. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. In particular, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first hypothesis favors the firing rate coding theory, the second emphasizes on the importance of temporal precision in neural activity. Since then, numerous studies have demonstrated the emergence of synchronous activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using precise spike timings~\citep{Perrinet2002,perrinet_coding_2004, gollisch_rapid_2008}, or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatiotemporal spiking motifs in biological spike trains. In neural models, the definition of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows the efficient detection of these spatiotemporal motifs embedded in the spike train. These spatiotemporal motifs present in neural activity may form useful representations to perform computations for various cognitive tasks using the synchrony of spikes reaching the soma of a neuron. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike at the soma of the afferent neuron and its arrival at the soma of the efferent neuron. Importantly, due to the variety of weights and delays within a population, representations using polychronous groups have in theory a much higher information capacity than a firing rate-based approach.

>>>>

The precise timing of spikes from afferent neurons determines the response of a biological neuron. This temporal aspect of neural activity is essential for understanding information processing in neurobiology, and is particularly applicable to the output of neuromorphic hardware such as event-based cameras. However, most artificial neural models ignore this important temporal dimension of neural activity. Here, we develop a model for efficiently detecting temporal spiking motifs based on a layer of spiking neurons with heterogeneous synaptic delays. The model exploits the property that the diversity of synaptic delays on the dendritic tree allows specific arrangements of temporally precise synaptic inputs to synchronize as they reach the basal dendritic tree. We show that this process can be formalized as a time-invariant logistic regression that can be trained on labeled data. We demonstrate its application to naturalistic videos transformed into event streams, similar to the output of the biological retina or event-based cameras. To characterize the robustness of the model in detecting visual motion, we test its efficiency by pruning some weights and show that it is still efficient at very low workloads. In conclusion, by providing an end-to-end, event-driven computational building block, the learning of heterogeneous delays could improve the performance of future spiking neural network (SNN) algorithms, and in particular their applications in neuromorphic chips.

@@

\keywords{\Keywords}
%
\maketitle
%
\section{Introduction}
%%%-----------------------------------------------------------------
\label{sec:intro}
%
The human brain has the remarkable capacity to react efficiently at any given time while consuming a reasonable amount of energy, in the order of 20 watts. This system is the result of millions of years of natural selection, and a striking difference with artificial neural networks is the representation that both use. For instance, the convolutional neural networks (CNNs) which are used for processing images represent the flow of information from one layer to another using tensors of floats, storing visual information densely across the visual topography, with different translation-invariant properties represented in different channels. CNNs have achieved state-of-the-art performance for some computer vision tasks, such as image recognition. These networks are also known to mimic several properties of biological systems, such that each can be assigned a ``brain score''~\citep{schrimpf_brain-score_2020}. However, this score does not take into account key aspects of the efficiency of biological systems, such as inference speed or energy consumption, which is about hundreds of watts on a standard GPU. In the vast majority of biological neural networks, on the other hand, information is represented as \emph{spikes}, prototypical all-or-nothing (binary) events whose only parameters are their timing and the address of the neuron that fired the spike~\citep{paugam-moisy_computing_2012}. Spiking neural networks (SNNs), known as the third generation of artificial neural networks, incorporate this temporal dimension into the way they perform their computations. One example of a SNN is the SpikeNet algorithm, which takes a purely temporal approach by encoding information using at most one spike per neuron~\citep{delorme_spikenet_1999}. Alternatively, the SNN implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} that uses the exact timing of spikes and learns the structure of the network by minimizing a specifically defined cost function. This was recently extended using the surrogate gradient method which is now widely used in attempts to transfer the performance of CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs still lags behind that of networks using an analog, firing rate-based representation. The question of the advantage of using spikes in machine learning and computer vision remains open.

In a recent review, we reported previous theoretical and  experimental evidence for the use of precise spiking motifs in biological neural networks~\cite{grimaldi_precise_2023}. In particular, \citet{abeles_role_1982} asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs. While the first hypothesis favors the firing rate coding theory, the second emphasizes on the importance of temporal precision in neural activity. Since then, numerous studies have demonstrated the emergence of synchronous activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding using precise spike timings~\citep{Perrinet2002,perrinet_coding_2004, gollisch_rapid_2008}, or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatiotemporal spiking motifs in biological spike trains. In neural models, the definition of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows the efficient detection of these spatiotemporal motifs embedded in the spike train. These spatiotemporal motifs present in neural activity may form useful representations to perform computations for various cognitive tasks using the synchrony of spikes reaching the soma of a neuron. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a spiking motif defined by a subset of neurons with different, but precise, relative spiking delays. This delay between a pair of connected neurons is defined as the time between the emission of a spike at the soma of the afferent neuron and its arrival at the soma of the efferent neuron. Importantly, due to the variety of weights and delays within a population, representations using polychronous groups have in theory a much higher information capacity than a firing rate-based approach.
